Red Wine Exploratory Analysis by Nathan Myers 
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that used for data analysis 

library(ggplot2)
library(dplyr)
library(gridExtra)
library(tidyr)
```

````{r echo=FALSE, cache=TRUE, Load_the_Data}
# Load the Data
wines <- read.csv("wineQualityReds.csv", row.names = 1, stringsAsFactors = FALSE)

wines$quality <- factor(wines$quality, ordered = TRUE)
```
#Introduction

This project uses R and data analysis techniques to explore a dataset
regarding the quality of certain redwines. A modest report on the data 
can be found at Elsevier, in addition to a short description 
of the individual variables.

The Dataset itself contains several physicochemical measures and attributes of
red vairiants of the Portugese wine "Vnho Verde" made and classified by wine
experts. 

# Univariate Plots
Taking a first look at the Data. 

```{r echo=FALSE, Glimpse}
glimpse(wines)
```

There are  `r ncol(wines)` variables and `r nrow(wines)` observations. 
All variables are numerical except for the quality score which is represented 
as a ordered factor.

## Quality

```{r echo=FALSE, Quality}
summary(wines$quality)
ggplot(data = wines, aes(x = quality))+
  geom_bar() 
``` 
The distribution of quality appears to be normal with many wines at average 
quality between 4 - 5 and a few wines at either end. Interestingly there are no
wines with a rating below 3 or higher than 8. 

##Fixed Acidity 

```{r echo=FALSE, fixed_acidity}
# Some functions to make it faster the univariate plotting
make_histogram <- function(variable, ...){
  ggplot(data = wines, aes_q(as.name(variable)))+
    geom_histogram(...)
}

print_summary <- function(variable){
  print(summary(wines[[variable]]))
}

plot_and_summary <- function(variable, ...){
  print_summary(variable)
  make_histogram(variable, ...)
}

plot_and_summary("fixed.acidity", binwidth = 0.2)+
  scale_x_continuous(breaks = 4:16)

```
Most of the wines have a fixed acidity between 7.1 and 9.20 and a median of 7.90
$g/dm^3.The distribution is slightly skewed to the right. Also notable are the 
several outliers. 

## Volatile acidity

```{r echo=FALSE, volatile.acidity}
plot_and_summary("volatile.acidity", binwidth = 0.02)+
  scale_x_continuous(breaks = seq(0, 1.6, 0.1))

```

The distribution of volatile acidity is non-symmetric and bimodal with two peaks
at 0.4 and 0.6. The median value is 0.52. Most observations within the data set
fall in the range 0.39 - 0.64.

## Citric acid

```{r echo=FALSE, citric.acid}
plot_and_summary("citric.acid", binwidth = 0.03)+
  scale_x_continuous(breaks = seq(0, 1, 0.125))

```
## Residual sugar

```{r echo=FALSE, residual.sugar}
plot_and_summary("residual.sugar", binwidth = 0.2)+
  scale_x_continuous(breaks = 0:16)

#limiting range
make_histogram("residual.sugar", binwidth = 0.2)+
  scale_x_continuous(breaks = 0:16)+
  coord_cartesian(xlim = c(0, 7))

```
The distribution of residual sugar has a median value of 2.2 $g/dm^3$. 
and skewed to the right with a long tail.

## Chlorides


```{r echo=FALSE, chlorides}
plot_and_summary("chlorides", binwidth = 0.01)+
  scale_x_continuous(breaks = seq(0, 0.6, 0.1))

#limiting range
make_histogram("chlorides", binwidth = 0.01)+
  scale_x_continuous(breaks = seq(0, 0.2, 0.05))+
  coord_cartesian(xlim = c(0, 0.2))

```

The amount of chlorides in the wines has a median value of 0.079 $g/dm^3$. 
The distribution appears normal around its main peak but has an unusually 
long right tail, with small counts of wines with values until 0.611 $g/dm^3$

## Free sulfur dioxide

```{r echo=FALSE, free.sulfur.dioxide}
plot_and_summary("free.sulfur.dioxide", binwidth = 2)+
  scale_x_continuous(breaks = seq(0, 70, 5))

```
The distribution of free sulfur dioxide has a median value of 14 $g/dm^3$, it 
is right skewed and a long tail. 

## Total sulfur dioxide


```{r echo=FALSE, total.sulfur.dioxide}
plot_and_summary("total.sulfur.dioxide", binwidth = 5)+
  scale_x_continuous(breaks = seq(0, 300, 25))

```

The distribution of total sulfur dioxide is right skewed. It has a median value 
of 38 $mg/dm^3$. On the right tail we can see a local maximum near 80. 
There's a gap between 165 and 278 with only two wines with a concentration 
greater than or equal to 278, with significant outliers betweem the 280 - 290 
mark. 

## Density


```{r echo=FALSE, density}
plot_and_summary("density", binwidth = 0.0005)+
  scale_x_continuous(breaks = seq(0.990, 1.005, 0.0025))

```

The density of wines varies , with most of the values between
0.9956 and 0.9967. The distribution is roughly symmetric and has median 
value of 0.9968 $g/cm^3$. The density if close to the density of 
water (1 $g/cm^3$ at 4 $^\circ C$).

## pH

```{r echo=FALSE, pH}
plot_and_summary("pH", binwidth = 0.02)+
  scale_x_continuous(breaks = seq(0, 4.5, 0.1))

```

All wines have low pH due to the fermantation process, making them acidic. This 
plot is the ambigious so far as the distribution could be either symmetrical or 
bimodal depending on the interpretation. There seems to be a local maximum at 
about and again at 3.35.  The median value is 3.31, and most 
ines have a pH between 3.21 and 3.4.

## Sulphates

```{r echo=FALSE, sulphates}
plot_and_summary("sulphates", binwidth = 0.05)+
  scale_x_continuous(breaks = seq(0, 2, 0.25))

```

The distribution of sulphates is slightly skewed to the right. Some outliers on 
the right tail have around 2 g/dm^3 of sulphates.The median value of sulphates 
is 0.62 and most wines have a concentration between 0.55 and 0.73.

## Alcohol

```{r echo=FALSE, alcohol}
plot_and_summary("alcohol", binwidth = 0.2)+
  scale_x_continuous(breaks = seq(8, 15, 1))

```

The alcohol concentration distribution is right skewed. The range of about 7,
with a min of 8.4 is exactly what we would expect for wine. The highest peak of 
the distribution is at 9.5 % alcohol and the median value is 10.20%. The maximum 
amount of alcohol present in the dataset is 14.90.

# Univariate Analysis

The dataset has 12 variables regarding 1599 observations. Each observation 
corresponds to a specific sample of red wine. 11 variables correspond to 
the result of a physicochemical test and one variable (`quality`) corresponds to 
the result of a sensory panel rating by experts

The main feature of interest is the correlation between various empirical
measurements and quality. I am also interested in why many of the distributions
have outliers.

There were no unusual distributions, missing values or need to adjust 
the data. The dataset presented clean and perfect for analysis.


# Bivariate Plots Section

```{r echo=FALSE}
make_box_with_dots <- function(variable){
  ggplot(data = wines, aes_q(x = ~quality, y = as.name(variable)))+
    geom_boxplot()+
    geom_jitter(width = 0.7, alpha = 1/10)+
    geom_line(aes(group = 1),
                  stat = "summary",
                  fun.y = median,
                  color = "#E74C3C",
                  size = 1,
                  alpha = 0.8)
}

cond_median <- function(variable){
  print(paste("Median of", variable, "by quality:"))
  print(by(wines[[variable]], wines$quality, median))
}

boxplot_and_median <- function(variable){
  cond_median(variable)
  make_box_with_dots(variable)
}
```

## Fixed Acidity vs. Quality


```{r echo=FALSE}
boxplot_and_median("fixed.acidity")

```

We see a slight upwards trend between quality and fixed acidity. However, 
the extreme quality classes (3 and 8) have less occurances, which may skew the 
median. There is also a slight aciddity drop from 7 to 8 in quality class. 
Additionally, we see a big dispersion of acidity values across each quality 
scale. This may be a indicator that the quality cannot be predicted based soley
on acidity and is the result of a combination of variables.

## Volatile Acidity vs. Quality

```{r echo=FALSE}
boxplot_and_median("volatile.acidity")

```

Compared to the Acidity plot we can see a more obvious trend. Lower volatile acidity seems to 
mean higher wine quality.

## Citric Acid vs. Quality

```{r echo=FALSE}
boxplot_and_median("citric.acid")

```

Higher citric acid content is strongly correlated with a higher quality wine. 
The citric acid is always in low concentrations and in the univariate plots we 
saw that the distribution peaked at the zero value.

What proportion of wines has zero citric acid. 

```{r echo=FALSE}
# How many have zero citric acid
# Total Percentage
sum(wines$citric.acid == 0) / nrow(wines)
```

For each class the proportions are:

```{r echo=FALSE}
# Grouped Percentage
group_by(wines, quality) %>%
  summarise(n_zero = sum(citric.acid == 0)/n())
```

We see a decreasing proportion of wines with zero citric acid on the higher 
quality classes.

So, this reinforces the first impression that the higher citric acid 
concentration relates to higher quality wines.

## Residual Sugar vs. Quality

```{r echo=FALSE, warning=FALSE}
boxplot_and_median("residual.sugar")

#removing the top 10% to be able to have a better look
boxplot_and_median("residual.sugar")+
  ylim(NA, quantile(wines$residual.sugar, 0.90))

```

Residual sugar has a low impact in the quality of the wine.

## Chlorides vs. Quality

```{r echo=FALSE, warning=FALSE}
boxplot_and_median("chlorides")

boxplot_and_median("chlorides")+
  ylim(NA, quantile(wines$chlorides, 0.90))

```

There is a slight relationship between lower chloride levels and quality.

## Free sulfur dioxide vs. Quality


```{r echo=FALSE}
boxplot_and_median("free.sulfur.dioxide")

```

The classes in the center of the distributions have a higher measurement 
of free sulfur dioxide than wines of high and low quality.

According to the dataset description, when free SO2 is lower than 50 ppm, 
roughly, ~ 50 mg/L, it is undetectable. The following plot shows that 
very few wines are above this threshold, leading us to believe  the 
variations are not related to levels free SO2, but the non balanced distribution 
of wines across the quality classes.

```{r echo=FALSE}
make_box_with_dots("free.sulfur.dioxide")+
  geom_hline(yintercept = 50, color = "#F1C40F", linetype = 2, size = 1.5)

```



## Total sulfur dioxide  vs. Quality

```{r echo = FALSE, warning=FALSE}
boxplot_and_median("total.sulfur.dioxide")+
  ylim(NA, 200)
```

Almost identical relationship as free sulfur dioxide, concentrated in the center
and then distributed low and high. 

## Density vs. Quality

```{r echo=FALSE}
boxplot_and_median("density")

```

Low density is significantly correlated with higher quality. The dataset describes 
the two most important factors of denisty as being alcohol and sugar content. 

## pH vs. Quality

```{r echo=FALSE}
boxplot_and_median("pH")

```

There is a slight relationship between acidity (low ph level) and High quality. 
Does this mean quality wine should taste acidic? heck correlations between 
pH and the acidity levels.

## Sulphates vs. Quality

```{r echo=FALSE, warning=FALSE}
boxplot_and_median("sulphates")

boxplot_and_median("sulphates")+
  ylim(NA, quantile(wines$sulphates, 0.90))

```

Strong relationship between high sulphate levels and high quality wine.

## Alcohol vs. Quality

```{r echo=FALSE}
boxplot_and_median("alcohol")

```

While average quality wine has slightly less alcohol than normal, above average 
quality wine has higher alcohol content.

## Acidity and pH

```{r echo=FALSE}
ggplot(wines, aes(x = pH, y = fixed.acidity))+
  geom_point(alpha = 1/3)+
  scale_y_log10()

```

This plot was done as a sort of control, as we would expect increases with the 
lower amount of acids. Fixed acidity accounts for most of the acid present in 
the wine.

```{r echo= FALSE}
ggplot(wines, aes(x = pH, y = citric.acid))+
  geom_point(alpha = 1/3)+
  scale_y_log10()

```

A similar relation is seen with the citric acid, but the relationsihp is not 
nearly as strong. This can mostly be attributed to the fact citric acid is only 
a small percentage of the acidic content. 

```{r echo=FALSE}
ggplot(wines, aes(x = pH, y = volatile.acidity))+
  geom_point(alpha = 1/3)+
  scale_y_log10()

```

The volatile acidity seems to have either no relation with the pH or a 
slight positive correlation.

Correlation coefficient:

```{r echo=FALSE}
cor.test(~ pH + log10(volatile.acidity), data = wines)

```

The correlation coefficient shows a weak positive correlation of volatile acidity
with the pH. Maybe when the volatile acids are present in higher concentration, 
the concentration of the remaining acids is lower and that contributes to the
increase of pH.

```{r echo=FALSE}
ggplot(wines, aes(x = volatile.acidity, y = fixed.acidity))+
  geom_point(alpha = 1/2)

cor.test(~ fixed.acidity + volatile.acidity, data = wines)

```


## Sulphates and sulfur oxide

Sulphate is an additive which can contribute to sulfur dioxide gas levels.

```{r echo=FALSE}
ggplot(wines, aes(x = sulphates, y = total.sulfur.dioxide))+
  geom_point(alpha = 1/5)

ggplot(wines, aes(x = sulphates, y = free.sulfur.dioxide))+
  geom_point(alpha = 1/5)

cor.test(~ total.sulfur.dioxide + sulphates, wines)
cor.test(~ free.sulfur.dioxide + sulphates, wines)

```

There is a very weak relationship between sulphat levels and sulphates.



## Correlations with quality

```{r echo=FALSE}
cor(x = wines[1:11],
    y = as.numeric(wines$quality),
    method = "spearman")

```

# Bivariate Analysis

There is a strong relationship between  wine quality and volatile acidity,
citric acid, sulphates and alcohol content. The correlation coefficients 
demonstrate relationships with the other variables.

Wines with average quality rankings, defined as between 5 and 6, have a higher
instance of free and total sulfur dioxide than wines at the extremes. 


I also looked at the relationship between ph, acidity levels. The correlation 
coefficients show that the variable with the strongest relationship with quality 
is the wine's alcohol content.

# Multivariate Plots Section

## Correlation Matrix

Constructing a correlation matrix:

```{r echo=FALSE}
# http://www.r-bloggers.com/using-r-correlation-heatmap-with-ggplot2/



correlations <- as.data.frame(cor(data.frame(wines[1:11],
                                  quality_num = as.numeric(wines$quality)),
                                  method = "spearman")) %>%
                add_rownames(var = "Var1") %>% # need row names for the gathering
                gather(key = "Var2",
                      value = "correlation",
                      -Var1)

ggplot(data = correlations, aes(x = Var1, y = Var2))+
  geom_tile(aes(fill = correlation))+
  geom_text(aes(label = round(correlation, 3)), size = 2)+
  scale_fill_gradient2(low = "#E74C3C", mid = "white", high="#2471A3")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, hjust = 1))


```

## Alcohol, volatile acidity and quality

Quality strongly correlates with alcohol and volatile acidity. 

```{r echo=FALSE}
scatter_with_quality_color <- function(x, y){
  ggplot(wines, aes_q(x = as.name(x), y = as.name(y), color = ~ quality))+
    geom_point(alpha = 0.8, size = 2)+
    scale_color_brewer(type = "div", palette = "RdYlGn")
}

scatter_with_quality_color("alcohol", "volatile.acidity")

```

The plot shows low quality wines have a low alcohol content paired with high 
volatile acidity. However mid-range wines can be found all over the 
distribution.

## Acidity, pH, quality

```{r echo=FALSE}
scatter_with_quality_color("fixed.acidity", "pH")+
  scale_x_log10()

```

There is no pattern or relationship between quality and fixed acidity. 

## Citric acid, alcohol, quality

```{r echo=FALSE}
scatter_with_quality_color("alcohol", "citric.acid")


```

There is definitely a relationship between higher wine quality and  an increase 
in citric acid and alcohol. However, there are also wines with average quality
ratings with a wide range of citric acid levels and low alcohol content.


## Alcohol and Sulphates

```{r echo=FALSE, warning=FALSE}
scatter_with_quality_color(x = "sulphates", y = "alcohol")

scatter_with_quality_color(x = "sulphates", y = "alcohol")+
  xlim(NA, 1.0)

```

Alcohol and sulphates are positively correlated. Additionaly, higher alcohol 
content combined with higher levels of sulphate results in higher quality wines.

```{r echo=FALSE}
cor.test(~ alcohol + sulphates, data = wines)

```



# Multivariate Analysis


Emphasis was primarily focused on variables with strong relationships. 

We have seen how alcohol and volatile acidity relate with quality. Higher 
alcohol and lower acidity give in general better quality wines.

Higher amounts of citric acid combined with higher alcohol content create the 
best wines.

Also with sulphates we see a similar trend, better quality wines have higher 
levels of alcohol and sulphates. 

------

# Final Plots and Summary

### Acetic Acd vs Quality
`### Plot One
```{r echo=FALSE, Plot_One, fig.width=14, fig.height=10}
make_box_with_dots("volatile.acidity")+
  xlab("Sensory Quality Rating")+
  ylab("Volatile Acidity (acetic acid - g / dm^3)")+
  ggtitle("Acetic acid vs quality")+
  theme(text = element_text(size = 20))
```

#### Description 

This plot shows us the distribution of volatile acidity across different quality
rankings. The scatterplot shows us the distribution of wines while the box plot
displays the quantile boundaries and median values. We can also observe most 
wines have a rating between 5-6, and there are more wines ranked at the higher
end of the scale, 7-8, than at the lower end. The red line connects the median
values and helps us better visualize the inverse relationship between volatile 
acidity and wine quality.


### Alcohol VS Quality

```{r echo=FALSE, Plot_Two, fig.width=14, fig.height=10}
make_box_with_dots("alcohol")+
  xlab("Sensory Quality Rating")+
  ylab("Alcohol (% by Volume)")+
  ggtitle("Alcohol VS Quality")+
  theme(text = element_text(size = 20))
```

#### Description

We can see a big impact of alcohol level on the quality of wines. For the quality 
classes 3 to 5, the effect is limited. The quality is probably being steered by 
another variable, but from the quality rating 5 to 8, we see a sharp increase in 
the alcohol content. The general trend is that Wines with higher alcohol content
are rated higher in quality.

There is a large impact on the level of alcohol and the quality of wines one the
higher end. Interestingly, there is a less dramatic effect for average and lower
quality wines. Howeverm the general trend is wines with a higher alcohol content
rate higher in quality. 
### Plot Three
```{r echo=FALSE, Plot_Three, fig.width=14, fig.height=10}
scatter_with_quality_color(x = "alcohol", y = "volatile.acidity")+
  xlab("Alcohol (% by Volume)")+
  ylab("Volatile Acidity (acetic acid - g / dm^3)")+
  ggtitle("Quality VS Volatile Acidity and Alcohol")+
  labs(color = "Quality Rating")+
  theme(text = element_text(size = 20))


```

#### Description Three

This plot shows the combined effect of volatile acidity and alcohol on wine 
quality. Wines with both high volatile acidity and low alcohol content have a 
lower quality rating and wines with low volatile acididty and low alcohol 
content have average ratings. Finally, wines with low volatile acididty and 
high alcohol content have the highest ratings. 

# Conclusion & Reflection

The Project was a compelling opportunity to tie together all of my knowledge of 
R and learn the nuances of exploring a dataset. The dataset itself was assembled 
specifically for data analysis and machine learning techniques, so it was already 
very well organized and not missing and data points. 

One of the most difficult challenges with working this kind of dataset was 
choosing the direction to steer the analysis in. Fortunately, the dataset 
description points to some variables of interest already, such as citric acid 
adding ‘freshness to wine’ or acetic acid adding a vinegary taste. I think this 
is an example of the significance of domain specific knowledge and how it and
can help guide our analysis. 

The second challenge i faced was explicating the meaning of a few of the 
multivariate plots. Adding a third dimension made it more difficult to view and 
interpret trends. Instead of relying on a line, you have to look closely at the 
changing colors. The correlation matrix was also helpful in finding variables 
with the closest relationships and to explore patterns. 

As a next step I think it would be interesting to add an analysis of white wines
and make comparisons. We could also, build a model to predict wine quality.
